---
title: "WeinR: Long-Read Sequencing Quality Control and Processing"
author:
  - name: Sarah Bowker
  - name: Owen Coelho
  - name: Nava Meadi
  - name: Myles Taillefer
  - name: Ido Hatam
    email: ihatam@langara.ca
    affiliation: Langara College
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('WeinR')`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{WeinR: Long-Read Sequencing Quality Control and Processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center"
)
```

# Introduction

`r Biocpkg("WeinR")` is an R/Bioconductor package for quality control (QC) and
preprocessing of Oxford Nanopore Technology (ONT) long-read sequencing data.
It accepts FASTQ, compressed FASTQ (`.fastq.gz`), and BAM input files (BAM
files must contain base quality scores in the `QUAL` field).

WeinR produces:

- Per-read and file-level QC metrics: read lengths, Phred quality scores, GC
  content, N counts, N50, N90, and total yield
- Five publication-quality PNG plots per file: length distribution, quality
  distribution, GC content distribution, quality-vs-length scatter, and
  per-position quality profile
- Interactive HTML reports with searchable summary tables and embedded plots
- Optionally, processed reads after filtering, adapter trimming, and/or end
  trimming

## Comparison with Similar Tools

Several tools address long-read QC, but most operate outside the R/Bioconductor
ecosystem or cover a narrower scope:

- **NanoStat** and **NanoPlot** are widely used Python command-line tools.
  They require a separate Python environment and produce outputs that are not
  directly interoperable with Bioconductor objects.
- **MinIONQC** is an R package focused on read-length and quality summaries,
  but does not support adapter trimming, end trimming, or chimera detection.
- **`r Biocpkg("Rfastp")`** provides adapter trimming and QC primarily for
  short reads; its long-read support does not include ONT-specific metrics
  such as per-position quality profiles or N50/N90.

WeinR covers the full ONT QC-and-processing pipeline within a single
R/Bioconductor package, returning results in a consistent S4 object that
integrates directly with downstream Bioconductor workflows.

## Package Workflow

The standard WeinR workflow has two phases:

1. **QC phase** — `CreateReport()` reads raw input, computes all metrics,
   generates plots, and optionally renders an HTML report.

2. **Processing phase** *(optional)* — `ProcessReads()` takes the `LongReadQC`
   object returned by `CreateReport()` and applies configurable filtering,
   adapter trimming, and/or end trimming.

All results accumulate in a single `LongReadQC` S4 object that is returned
from each function and saved to disk.

# Installation

`WeinR` can be installed from Bioconductor using `r CRANpkg("BiocManager")`:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("WeinR")
```

To install the development version directly from GitHub:

```{r install-gh, eval=FALSE}
BiocManager::install("idohatam/WeinR")
```

Once installed, load the package:

```{r load}
library(WeinR)
```

# Example Data

This vignette uses a small FASTQ file bundled with the package under
`inst/extdata/`. It contains 300 synthetic reads with ONT-like properties:
a right-skewed length distribution (median ~3.9 kb, range 424--12,000 bp),
per-read GC content centred near 50%, and a 3' quality drop-off typical of
ONT chemistry (mean Q ~14). Using `system.file()` to locate bundled data
avoids any external downloads and ensures the vignette is fully self-contained.

```{r example-data}
fastq_file <- system.file("extdata", "sample.fastq", package = "WeinR")
if (!nzchar(fastq_file))
    stop("'sample.fastq' was not found in the WeinR package installation.")
```

# Basic Workflow: `CreateReport()`

`CreateReport()` is the primary entry point. Provide either a **vector of file
paths** or a **directory** (WeinR will discover `.fastq`, `.fastq.gz`, and
`.bam` files automatically). The function returns a `LongReadQC` object and
writes outputs under `WeinR_Outputs/` in the current working directory.

```{r run-qc}
qc_obj <- CreateReport(
  files         = fastq_file,
  MinNumReads   = 1L,
  report_name   = "example_qc",
  render_report = FALSE,   # set TRUE to also produce an HTML report
  force         = TRUE     # overwrite any existing outputs from prior runs
)

qc_obj
```

## The `LongReadQC` Object

All results are stored in the returned `LongReadQC` S4 object. The five slots
and their contents are:

| Slot | Contents |
|------|----------|
| `@files` | Character vector of input file paths processed |
| `@metrics` | Named list of per-read metric vectors/data frames, keyed by filename |
| `@plots` | Named list of saved plot file paths, keyed by filename |
| `@summary_metrics` | Data frame of file-level summary statistics |
| `@metadata` | List of processing summaries populated by `ProcessReads()` |

```{r inspect-files}
basename(qc_obj@files)
```

The per-file metrics available for the example file:

```{r inspect-metrics}
fname <- basename(fastq_file)
names(qc_obj@metrics[[fname]])
```

# Output Walkthrough

## Summary Metrics Table

`@summary_metrics` provides one row per input file:

```{r summary-table}
qc_obj@summary_metrics
```

| Column | Description |
|--------|-------------|
| **File** | Input filename |
| **Read Count** | Total number of reads in the file |
| **Yield (bp)** | Total bases sequenced across all reads |
| **N50** | Read length at which 50% of all sequenced bases are in reads of this length or longer |
| **N90** | Same as N50 but for 90% of bases; always ≤ N50 |
| **Average Per Read Q Score** | Mean Phred quality score averaged across all reads |
| **N Count** | Total number of ambiguous (`N`) bases across all reads |

## Per-Read Metrics

The `@metrics` slot stores the raw per-read vectors that feed the plots:

```{r per-read-metrics}
m <- qc_obj@metrics[[fname]]

summary(m$readLengths)          # read lengths (bp)
summary(unlist(m$meanprQscore)) # mean Phred Q-score per read
summary(unlist(m$prGCcontent))  # per-read GC content (fraction)
```

Per-position quality is stored as a data frame with one row per position
window and columns `position` (bp), `mean`, `median`, `q25`, and `q75`:

```{r per-pos-preview}
head(m$perPosQuality)
```

## Output Folder Structure

`CreateReport()` writes all outputs under `WeinR_Outputs/` in the working
directory, inside a timestamped run folder:

```
WeinR_Outputs/
└── run_YYYY-MM-DD_HHMMSS/
    ├── metrics/
    │   ├── sample.fastq/
    │   │   ├── readLengths.csv
    │   │   ├── meanprQscore.csv
    │   │   ├── prGCcontent.csv
    │   │   ├── Ncount.csv
    │   │   └── perPosQuality.csv
    │   └── example_qc_summary_metrics.csv
    ├── plots/
    │   └── sample/
    │       ├── length_hist.png
    │       ├── quality_hist.png
    │       ├── gc_hist.png
    │       ├── q_vs_length.png
    │       └── per_pos_q.png
    ├── reports/
    │   ├── example_qc.Rmd
    │   └── example_qc.html   # only produced when render_report = TRUE
    └── objects/
        └── qc.rds            # saved LongReadQC object
```

All CSV files can be imported directly for downstream analysis. To reload the
`LongReadQC` object in a future R session, use `readRDS()` with the path
printed by `CreateReport()` at runtime.

## Plots

Five PNG plots are generated per file and their paths stored in `@plots`.
The sub-sections below describe each plot and what to look for in real data.

### Read Length Distribution

A histogram of per-read lengths on a log~10~ x-axis. In a typical ONT run,
the bulk of reads will fall between 1 and 20 kb, with a right-skewed tail.
Very short reads concentrated at the left of the distribution may indicate
pore blockage, degraded input, or adapter dimers.

```{r plot-length, echo=FALSE, out.width="70%"}
knitr::include_graphics(qc_obj@plots[[fname]][["length_hist"]])
```

### Average Q-Score Distribution

A histogram of per-read mean Phred quality scores, with vertical lines for
the mean (dashed red) and median (solid blue). Current ONT chemistry typically
produces reads centred around Q15--Q20. A bimodal distribution can indicate
two populations with different pore states or chemistry.

```{r plot-quality, echo=FALSE, out.width="70%"}
knitr::include_graphics(qc_obj@plots[[fname]][["quality_hist"]])
```

### GC Content Distribution

A histogram of per-read GC fraction. The mode should correspond to the
expected GC content of the sequenced organism. Broad or heavily skewed
distributions may indicate contamination or adapter carry-through.

```{r plot-gc, echo=FALSE, out.width="70%"}
knitr::include_graphics(qc_obj@plots[[fname]][["gc_hist"]])
```

### Quality vs Read Length

A scatter plot of per-read mean Q-score against read length (log~10~ x-axis).
Points are rasterised for performance with large datasets. In ONT data, longer
reads often show modestly lower quality due to signal drift over extended pore
occupancy.

```{r plot-qvl, echo=FALSE, out.width="70%"}
knitr::include_graphics(qc_obj@plots[[fname]][["q_vs_length"]])
```

### Per-Position Quality Profile

A smoothed profile of Phred quality along the read coordinate. The solid line
is the median Q-score; the ribbon spans Q25--Q75. In real ONT runs, this plot
typically shows a sharp quality spike at the 5' end followed by a rapid drop,
a noisy but stable middle region, and a pronounced decline toward the 3' end.
A sharp dip near the 5' end can suggest adapter contamination or a poor run
start. The profile below is derived from the bundled synthetic dataset (reads
up to 12,000 bp); real ONT data with longer reads will produce a richer
profile with a more pronounced 5' spike and sharper 3' decline.

```{r plot-perpos, echo=FALSE, out.width="70%"}
knitr::include_graphics(qc_obj@plots[[fname]][["per_pos_q"]])
```

# Read Processing: `ProcessReads()`

After reviewing the QC output, use `ProcessReads()` to prepare reads for
downstream analysis. It takes the `LongReadQC` object from `CreateReport()`
and applies up to three steps in a fixed order:

1. **Quality / length / N filtering** (`filter = TRUE`)
2. **Adapter trimming** (`AdapterSeq = "..."`)
3. **Fixed end trimming** (`Start = n`, `End = n`)

Any combination of steps can be enabled; unused steps are skipped. Processed
reads are written to `WeinR_Outputs/Processed_Files/`.

## Filtering Only

```{r process-filter}
qc_filtered <- ProcessReads(
  qc_obj,
  filter      = TRUE,
  MinAvgQS    = 10,    # drop reads with mean Q-score < 10
  MinLength   = 150,   # drop reads shorter than 150 bp
  MaxNumberNs = 2,     # drop reads with > 2 N bases
  outpath       = "example_filtered",
  title         = "Filtered QC",
  render_report = FALSE,
  force         = TRUE,
  verbose       = FALSE
)

qc_filtered
```

## Full Processing Workflow

```{r process-full}
qc_processed <- ProcessReads(
  qc_obj,

  # Step 1: filtering
  filter      = TRUE,
  MinAvgQS    = 10,
  MinLength   = 150,
  MaxNumberNs = 2,

  # Step 2: adapter trimming
  # Replace with the adapter sequence for your sequencing kit
  AdapterSeq          = "AGATCGG",
  MaxMismatchEnd      = 3,
  MinOverlapEnd       = 20,
  MinInternalDistance = 100,
  MinFragmentLength   = 150,

  # Step 3: end trimming
  Start = 2,
  End   = 2,

  OutFileType       = "fastq",
  FinalSuffix       = "processed",
  KeepIntermediates = FALSE,

  outpath       = "example_processed",
  title         = "Full Processing Workflow",
  render_report = FALSE,
  force         = TRUE,
  verbose       = FALSE
)

qc_processed
```

## Inspecting Processing Metadata

`ProcessReads()` records reads-before and reads-after counts for every
step in `@metadata`:

```{r metadata}
meta <- qc_processed@metadata[[fname]]
names(meta)

meta$filter_summary[c("reads_before", "reads_after")]
```

## `ProcessReads()` Parameter Reference

| Group | Parameter | Default | Description |
|-------|-----------|---------|-------------|
| **Filtering** | `filter` | `FALSE` | Enable quality/length/N filtering |
| | `MinAvgQS` | `20` | Minimum mean Phred Q-score per read |
| | `MinLength` | `200` | Minimum read length (bp) |
| | `MaxNumberNs` | `2` | Maximum ambiguous bases per read |
| **Adapter trimming** | `AdapterSeq` | `NULL` | Adapter sequence; `NULL` skips this step |
| | `MaxMismatchEnd` | `3` | Allowed mismatches for end-adapter matching |
| | `MinOverlapEnd` | `20` | Distance from read end for end-adapter classification |
| | `MinInternalDistance` | `100` | Distance from ends for chimera detection |
| | `MinFragmentLength` | `200` | Minimum fragment length after chimera splitting |
| **End trimming** | `Start` | `NULL` | Bases to remove from 5' end; `NULL` skips |
| | `End` | `NULL` | Bases to remove from 3' end; `NULL` skips |
| **Output** | `OutFileType` | `"fastq"` | One or more of `"fastq"`, `"bam"`, `"fasta"` |
| | `FinalSuffix` | `"processed"` | Suffix appended to output filenames |
| | `KeepIntermediates` | `FALSE` | Retain per-step intermediate FASTQ files |

# Session Info

```{r session-info}
sessionInfo()
```
