---
title: "WeinR-intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WeinR-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(WeinR)
```


## Introduction
`WeinR` provides tools for long-read quality control, preprocessing, and report generation.
The package is designed for workflows involving FASTQ, FASTQ.GZ, and BAM input files,
with support for computing QC metrics, producing plots, processing reads, and generating
HTML reports.

This vignette demonstrates two common entry points:

- `CreateReport()` for a lightweight quality-control workflow
- `ProcessReads()` for filtering, adapter trimming, end trimming, and reporting

## Create example input data
```{r create-data}
tmp_dir <- tempfile("weinr_vignette_")
dir.create(tmp_dir)

fastq_file <- file.path(tmp_dir, "example.fastq")

writeLines(c(
  "@read1",
  "ACTGACTGACTG",
  "+",
  "IIIIIIIIIIII",
  "@read2",
  "ACTGNNNNACTG",
  "+",
  "####IIII####"
), fastq_file)

fastq_file
```

## Run QC with `CreateReport()`
The `CreateReport()` function runs quality control on one or more input files and returns
a `LongReadQC` object containing summary metrics, plots, and metadata.

```{r run-qc}
qc_obj <- CreateReport(
  files = tmp_dir,
  MinNumReads = 1,
  report_name = "example_report",
  render_report = FALSE,
  force = TRUE
)

qc_obj
```

## Inspecting the QC object
A `LongReadQC` object stores information about the files, metrics, plots, summary tables,
and metadata generated during QC.

```{r inspect-qc}
slotNames(qc_obj)

qc_obj@summary_metrics

names(qc_obj@metrics)

qc_obj@plots
```

## Process reads with `ProcessReads()`
After generating QC results, the `ProcessReads()` function can be used to filter and
process reads. This workflow can include filtering, adapter removal, and end trimming.

Below is a simple example using filtering only.

```{r process-filter}
qc_filtered <- ProcessReads(
  qc_obj,
  filter = TRUE,
  MinAvgQS = 20,
  MinLength = 4,
  MaxNumberNs = 2,
  outpath = "qc_filtered",
  title = "Filtered QC Report",
  render_report = FALSE,
  force = TRUE,
  verbose = TRUE
)

qc_filtered
```

## Run a fuller workflow
`ProcessReads()` can combine multiple steps in a single workflow. For example, filtering,
adapter trimming, and end trimming can all be performed in one call.

```{r process-full}
qc_full <- ProcessReads(
  qc_obj,
  filter = TRUE,
  MinAvgQS = 20,
  MinLength = 4,
  MaxNumberNs = 2,
  AdapterSeq = "AGAT",
  Start = 1,
  End = 1,
  OutFileType = "fastq",
  outpath = "qc_full",
  title = "Full Processing Workflow",
  render_report = FALSE,
  force = TRUE,
  verbose = TRUE,
  KeepIntermediates = FALSE,
  FinalSuffix = "processed"
)

qc_full
```

## Output files
`WeinR` writes outputs under WeinR_Outputs, including generated reports and saved QC objects.
Processed reads are written by the workflow steps used during processing.

## Session info

```{r session-info}
sessionInfo()
```

## Summary

This vignette demonstrated a basic `WeinR` workflow, including quality control,
inspection of QC metrics, and read processing using `CreateReport()` and `ProcessReads()`.
